Project Membach

### 2020-05-29 project.Rmd
working on draft

setwd("/media/data/Documents/Uni/spatiotemp/villalobos/project/")
if (!exists("domain")) {
  # UCCS
  domain <- c(-350000,350000); pattern <- "UCCS.*\\.csv"; name <- "UCCS"; description <- "Uccle, Bruxelles"; proj_file <- 'proj.uccs.RData'
  # MEMS
  domain <- c(-7200,7200); pattern <- "MEM.*\\.csv"; name <- "MEM"; description <- "Membach"; proj_file <- 'proj.mems.RData'
}
if (exists("proj_file")) {
  load(paste0("/media/data/Documents/Uni/spatiotemp/villalobos/project/", proj_file))
  # save.image(paste0("/media/data/Documents/Uni/spatiotemp/villalobos/project/", proj_file))
}

@@

# Data preparation
```{r dataset description, echo=FALSE}
long_description <- description %s+% " (" %s+% name %s+% " station)"
long_description
"Domain values: " %s+% domain[1] %s+% " to " %s+% domain[2] %s+% " nm"
"Data from " %s+% format.POSIXct(min(seis$utc_second), "%Y-%m-%d %H:%M:%S") %s+% " to " %s+% format.POSIXct(max(seis$utc_second), "%Y-%m-%d %H:%M:%S")
```

```{r cleanup, echo=FALSE}
out_domain <- (seis$min < domain[1] | seis$max > domain[2])
invalid_values <- (seis$min == -1 & seis$max == 1) | (seis$min == 0 & seis$max == 0) | out_domain
"Removing " %s+% sum(out_domain) %s+% " values out of domain, " %s+% (sum(invalid_values) - sum(out_domain)) %s+% " explicit NA"
seis <-
  seis %>%
  mutate(min=replace(min, invalid_values, NA), max=replace(max, invalid_values, NA)) %>%
  drop_na()
# %>% mutate(belgian_second=with_tz(utc_second, "Europe/Brussels"))
remove(out_domain, invalid_values) # free ~350 Mb from RAM
```

@@

seis.present <- length(seis$utc_second)
seis.total <- length(seis.complete$utc_second)
seis.missing <- seis.total - seis.present
  seis.present %s+% " present values, " %s+% seis.missing %s+% " missing points (" %s+% print_dec(seis.present / seis.total * 100, 2) %s+% "% / " %s+% print_dec(seis.missing / seis.total * 100, 2) %s+% "%)"

@@

block_start <- NULL
block_end <- NULL
blocks <- data.frame(start=integer(), end=integer())
for (i in 1:length(seis.missing.ids)) {
  if (is.null(block_start)) {
    block_start <- seis.missing.ids[i]
    block_end <- seis.missing.ids[i]
  } else if ((block_end + 1) == seis.missing.ids[i]) {
    block_end <- seis.missing.ids[i]
  } else {
    blocks[nrow(blocks) + 1, ] <- c(block_start, block_end)
    block_start <- seis.missing.ids[i]
    block_end <- seis.missing.ids[i]
  }
}
blocks[nrow(blocks) + 1, ] <- c(block_start, block_end)
blocks$interval <- hours(blocks$end - blocks$start + 1)
blocks$start <- seis.first.second + hours(blocks$start)
blocks$end <- seis.first.second + hours(blocks$end)
plot(c(rbind(blocks$start, blocks$end, NA)), rep(hour(blocks$interval), each=3),
     col=ifelse(blocks$interval<3600, "darkgreen", "darkred"),
     main="Consecutive hours of missing values",
     ylab="Hours", xlab="Interval start",
     type="l", lwd=10,
     xaxt="n", xlim=c(min(seis$utc_second), max(seis$utc_second))
     )
axis.POSIXct(1, at=seq.POSIXt(from=min(seis$utc_second), to=max(seis$utc_second), by = "1 month"), format="%m-%y", cex=0.8)
text(x=blocks$start, y=hour(blocks$interval), labels = paste0(format(blocks$start, "%y-%m-%d %H:%M")), cex=0.9, pos=2)
text(x=blocks$end, y=hour(blocks$interval), labels = paste0(format(blocks$end, "%y-%m-%d %H:%M")), cex=0.9, pos=4)
text(x=blocks$start + make_difftime(hour=sapply(hour(blocks$interval), "/", y=2)), y=hour(blocks$interval), labels = paste0(hour(blocks$interval), "H"), cex=0.9, pos=3)
text(x=blocks$start + make_difftime(hour=sapply(hour(blocks$interval), "/", y=2)), y=hour(blocks$interval), labels = paste0(hour(blocks$interval), "H"), cex=0.9, pos=1)
remove(seis.complete, seis.missing.ids)

@@

```{r periodgram}
library(spec)
pgram <- spec.pgram(build_ts(seis.h, start = seis.train.start, end = seis.train.end), main="periodgram", detrend=F, demean=T, na.action=na.aggregate)
topfreq <- pgram$freq[order(pgram$spec, decreasing = T)][1:100]
topseas <- 1/topfreq
topseas.h <- topseas/3600

filter_freq <- (pgram$spec > (mean(pgram$spec) + sd(pgram$spec)*2)) 
top95pc <- pgram$spec[filter_freq]
top95freq <- pgram$freq[filter_freq][order(top95pc, decreasing = T)]
top95pc <- pgram$spec[filter_freq][order(top95pc, decreasing = T)]
cat("SD ratio\n")
(top95pc - mean(pgram$spec))/sd(pgram$spec)
top95seas.s <- 1/top95freq
cat("Raw frequency by unit\n")
top95seas.s
top95seas.h <- top95seas.s/3600
cat("Frequency by unit\n")
top95seas.h

ylims <- 10^par("usr")[3:4]
axis(side=3, at=top95freq, labels=as.integer(top95seas.h), padj = 1, las=2, cex.axis=0.8)
abline(h=mean(pgram$spec) + sd(pgram$spec)*c(.3,1:3), lty="dotted", col=c("orange", "darkgreen", "blue", "red"))
for (i in 1:length(top95freq)) {
  color <- ifelse(top95pc[i] > (mean(pgram$spec) + sd(pgram$spec)*3), "red", ifelse(top95pc[i] > (mean(pgram$spec) + sd(pgram$spec)*2), "blue", ifelse(top95pc[i] > (mean(pgram$spec) + sd(pgram$spec)*1), "darkgreen", "orange")))
  lines(x=c(top95freq[i], top95freq[i]), y=c(top95pc[i],ylims[2]), lty="dotted", col=color)
  points(x=top95freq[i], y=top95pc[i], col=color)
}
axis(side=4, at=mean(pgram$spec) + sd(pgram$spec) * c(.3, 1:3), labels=c(".3", "1", "2", "3"), cex.axis=0.8)
mtext(side=4, "spectrum sd ratio", line=1)
gc()
```

@@

```
Once notice the composition is additive (the fluctuations does not seem to grow with the trend)
```{r decompose}
dec.addi <- decompose(seasonality.aggregate(seis.h.train))
autoplot(dec.addi) + labs(title="Decomposition of additive time series", subtitle = paste0("Seasonality ", seasonality.period))
dec.multi <- decompose(log(seasonality.aggregate(seis.h.train)))
autoplot(dec.multi) + labs(title="Decomposition of multiplicative time series (log-scale additive)", subtitle = paste0("Seasonality ", seasonality.period))
plot_grid(
  autoplot(acf(dec.addi$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  autoplot(acf(dec.multi$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  nrow = 2)
plot_grid(
  autoplot(pacf(dec.addi$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  autoplot(pacf(dec.multi$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  nrow = 2)
```

>>>>

setwd("/media/data/Documents/Uni/spatiotemp/villalobos/project/")
if (exists("proj_file")) {
  load(paste0("/media/data/Documents/Uni/spatiotemp/villalobos/project/", proj_file))
  # save.image(paste0("/media/data/Documents/Uni/spatiotemp/villalobos/project/", proj_file))
}

@@

# Data preparation
```{r "dataset description", echo=FALSE}
long_description <- description %s+% " (" %s+% name %s+% " station)"
cat(long_description, "\n")
cat("Domain values:", domain[1], "to", domain[2], "nm\n")
cat("Data from", format.POSIXct(min(seis$utc_second), "%Y-%m-%d %H:%M:%S"), "to", format.POSIXct(max(seis$utc_second), "%Y-%m-%d %H:%M:%S"), "\n")
```

```{r cleanup, echo=FALSE}
out_domain <- (seis$min < domain[1] | seis$max > domain[2])
invalid_values <- (seis$min == -1 & seis$max == 1) | (seis$min == 0 & seis$max == 0) | out_domain
cat("Removing", sum(out_domain), "values out of domain,", (sum(invalid_values) - sum(out_domain)), "explicit NA\n")
seis <-
  seis %>%
  mutate(min=replace(min, invalid_values, NA), max=replace(max, invalid_values, NA)) %>%
  drop_na()
# %>% mutate(belgian_second=with_tz(utc_second, "Europe/Brussels"))
remove(out_domain, invalid_values) # free ~350 Mb from RAM
```

@@

seis.present <- length(seis$utc_second)
seis.total <- length(seis.complete$utc_second)
seis.missing <- seis.total - seis.present
cat(seis.present, "present values,", seis.missing, "missing points (", print_dec(seis.present / seis.total * 100, 2), "% /", print_dec(seis.missing / seis.total * 100, 2), "% )\n")

@@

block_start <- NULL
block_end <- NULL
blocks <- data.frame(start=integer(), end=integer())
for (i in 1:length(seis.missing.ids)) {
  if (is.null(block_start)) {
    block_start <- seis.missing.ids[i]
    block_end <- seis.missing.ids[i]
  } else if ((block_end + 1) == seis.missing.ids[i]) {
    block_end <- seis.missing.ids[i]
  } else {
    blocks[nrow(blocks) + 1, ] <- c(block_start, block_end)
    block_start <- seis.missing.ids[i]
    block_end <- seis.missing.ids[i]
  }
}
blocks[nrow(blocks) + 1, ] <- c(block_start, block_end)
blocks$interval <- hours(blocks$end - blocks$start + 1)
blocks$start <- seis.first.second + hours(blocks$start)
blocks$end <- seis.first.second + hours(blocks$end)
plot(c(rbind(blocks$start, blocks$end, NA)), rep(hour(blocks$interval), each=3),
     col=ifelse(blocks$interval<3600, "darkgreen", "darkred"),
     main="Consecutive hours of missing values",
     ylab="Hours", xlab="Interval start (mm-'yy)",
     type="l", lwd=10,
     xaxt="n", xlim=c(min(seis$utc_second), max(seis$utc_second))
     )
axis.POSIXct(1, at=seq.POSIXt(from=min(seis$utc_second), to=max(seis$utc_second), by = "1 month"), format="%m-'%y", cex=0.8)
text(x=blocks$start, y=hour(blocks$interval), labels = paste0(format(blocks$start, "%y-%m-%d %H:%M")), cex=0.9, pos=2)
text(x=blocks$end, y=hour(blocks$interval), labels = paste0(format(blocks$end, "%y-%m-%d %H:%M")), cex=0.9, pos=4)
text(x=blocks$start + make_difftime(hour=sapply(hour(blocks$interval), "/", y=2)), y=hour(blocks$interval), labels = paste0(hour(blocks$interval), "H"), cex=0.9, pos=3)
text(x=blocks$start + make_difftime(hour=sapply(hour(blocks$interval), "/", y=2)), y=hour(blocks$interval), labels = paste0(hour(blocks$interval), "H"), cex=0.9, pos=1)
remove(seis.complete, seis.missing.ids)

@@

```{r training data transformation plots}
month.starts <- seq.POSIXt(from=floor_date(seis.train.start, 'month'), to=ceiling_date(seis.train.end, 'month'), by = "1 month")
hours.from.train.start <- as.numeric(month.starts-seis.train.start)*24+168

autoplot(seis.h.train, type="l") +
  scale_x_continuous(breaks=hours.from.train.start/168, labels=format(month.starts, "%m-%y")) +
  ylab("mean hourly movement (nm)") + xlab(NULL) + ggtitle("Training set") + labs(caption = "Data in original scale")
gghistogram(seis.h.train, add.normal = T) + xlab("mean hourly movement (nm)") + labs(caption = "Data in original scale")
gg_qqplot(seis.h.train) + 
  labs(title="Q-Q plot for training dataset") + ylab("Sample quantiles") + labs(caption = "Data in original scale")

autoplot(log(seis.h.train), type="l") +
  scale_x_continuous(breaks=hours.from.train.start/168, labels=format(month.starts, "%m-%y")) +
  ylab("mean hourly movement (nm)") + xlab(NULL) + ggtitle("Training set") + labs(caption = "Data in log-scale")
gghistogram(log(seis.h.train), add.normal = T) + xlab("mean hourly movement (nm)") + labs(caption = "Data in log-scale")
gg_qqplot(log(seis.h.train)) + 
  labs(title="Q-Q plot for training dataset") + ylab("Sample quantiles") + labs(caption = "Data in log-scale")

train.bc.lambda <- BoxCox.lambda(seis.h.train)
train.bc <- BoxCox(seis.h.train, train.bc.lambda)
autoplot(train.bc, type="l") +
  scale_x_continuous(breaks=hours.from.train.start/168, labels=format(month.starts, "%m-%y")) +
  ylab("mean hourly movement (nm)") + xlab(NULL) + ggtitle("Training set") + labs(caption = paste0("Data Box-Cox transformed with lambda = ", train.bc.lambda))
gghistogram(train.bc, add.normal = T) + xlab("mean hourly movement (nm)") + labs(caption = paste0("Data Box-Cox transformed with lambda = ", train.bc.lambda))
gg_qqplot(train.bc) + 
  labs(title="Q-Q plot for training dataset") + ylab("Sample quantiles") + labs(caption = paste0("Data Box-Cox transformed with lambda = ", train.bc.lambda))
```

```{r periodgram}
library(spec)
pgram <- spec.pgram(build_ts(seis.h, start = seis.train.start, end = seis.train.end), main="Periodgram", detrend=F, demean=T, na.action=na.aggregate, plot=F, xlim=c(0,1/3600))
pgram.data <- data.frame(freq=pgram$freq, spec=pgram$spec)
ggplot(pgram.data) +
  geom_line(aes(freq, log(spec))) +
  xlab("frequency") + ylab("log(spectrum)") +
  labs(title = "Periodgram", caption = paste0("bandwidth = ", pgram$bandwidth, "\tdf = ", pgram$df)) ->
  periodgram

topfreq <- pgram$freq[order(pgram$spec, decreasing = T)][1:100]
topseas <- 1/topfreq
topseas.h <- topseas/3600

start_ratio <- 2
filter_freq <- (pgram$spec > (mean(pgram$spec) + sd(pgram$spec)*start_ratio)) 
top95pc <- pgram$spec[filter_freq]
top95freq <- pgram$freq[filter_freq][order(top95pc, decreasing = T)]
top95pc <- pgram$spec[filter_freq][order(top95pc, decreasing = T)]
cat("SD ratio\n")
(top95pc - mean(pgram$spec))/sd(pgram$spec)
top95seas.s <- 1/top95freq
cat("Raw frequency by unit\n")
top95seas.s
top95seas.h <- top95seas.s/3600
cat("Frequency by unit\n")
top95seas.h

ratio.color <- function(x) ifelse(x >= (mean(pgram$spec) + sd(pgram$spec)*3), "red", ifelse(x >= (mean(pgram$spec) + sd(pgram$spec)*2), "blue", ifelse(x >= (mean(pgram$spec) + sd(pgram$spec)*1), "darkgreen", "orange")))

library(ggrepel)
periodgram +
  scale_x_continuous(sec.axis = sec_axis(~1/./3600, breaks=c(1, 6*c(1,2,4,4*7)), labels = as.integer, name="hours")) +
  scale_y_continuous(sec.axis = sec_axis(~exp(.), breaks=mean(pgram$spec) + sd(pgram$spec) * c(1:3), labels=c("1", "2", "3")[1:3], name="spectrum/sd ratio")) +
  geom_hline(yintercept=log(mean(pgram$spec) + sd(pgram$spec) * c(1:3)), linetype="dotted", color = ratio.color(mean(pgram$spec) + sd(pgram$spec) * c(1:3))) +
  geom_point(data=data.frame(freq=top95freq, spec=top95pc), aes(freq, log(spec), color=ratio.color(spec)), shape="o", size=4, show.legend = F) +
  geom_label_repel(data=data.frame(freq=top95freq, spec=top95pc)[1:3,], aes(freq, log(spec), label=round(1/freq/3600), color=ratio.color(spec)), segment.color="grey", show.legend=F) +
  scale_color_manual(values = c("darkgreen", "blue", "red")[start_ratio:3])

@@

```
Once notice the composition is additive (the fluctuations does not seem to grow with the trend)
```{r decompose}
dec.addi <- decompose(seasonality.aggregate(seis.h.train))
autoplot(dec.addi) + labs(title="Decomposition of additive time series", subtitle = paste0("Seasonality ", seasonality.period))
dec.multi <- decompose(log(seasonality.aggregate(seis.h.train)))
autoplot(dec.multi) + labs(title="Decomposition of multiplicative time series (log-scale additive)", subtitle = paste0("Seasonality ", seasonality.period))
dec.multi.plain <- decompose(seasonality.aggregate(seis.h.train), type="multip")
autoplot(dec.multi.plain) + labs(title="Decomposition of multiplicative time series", subtitle = paste0("Seasonality ", seasonality.period))
plot_grid(
  autoplot(acf(dec.addi$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  autoplot(acf(dec.multi$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  autoplot(acf(dec.multi.plain$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  nrow = 3)
plot_grid(
  autoplot(pacf(dec.addi$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  autoplot(pacf(dec.multi$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  autoplot(pacf(dec.multi.plain$random, na.action=na.omit, plot=F)) + scale_x_continuous(),
  nrow = 3)
```

### 2020-05-29 project.mems.Rmd
working on draft

---
title: "Project Membach"
author: "Marco Ciotola"
date: "16/05/2020"
output:
  pdf_document:
    keep_tex: yes
    fig_width: 6.5
    fig_height: 4
#  html_document:
#    mathjax: local
#    self_contained: no
#    fig_width: 6.5
#    fig_height: 4
---

@@

```{r transformation}
plot(seis.h, xaxt="n", type="l")
axis.POSIXct(1, at=seq.POSIXt(from=seis.h.start, to=seis.h.end, by = "1 month"), format="%m-%y", cex=0.8)

hist(seis.h.train)
plot(log(seis.h.train))
hist(log(seis.h.train))
train.bc.lambda <- BoxCox.lambda(seis.h.train)
train.bc <- BoxCox(seis.h.train, train.bc.lambda)
plot(train.bc)
gghistogram(train.bc, add.normal = T)
gg_qqplot(seis.h.train) + 
        labs(title="Q-Q plot for training dataset") + ylab("Sample quantiles")
gg_qqplot(log(seis.h.train)) + 
        labs(title="Q-Q plot for log(training) dataset") + ylab("Sample quantiles")
gg_qqplot(train.bc) + 
        labs(title="Q-Q plot for BoxCox(training) dataset") + ylab("Sample quantiles")
```

```{r acf-pacf}
train.bc %>% ggtsdisplay(lag.max = 168*5)
train.bc %>% diff() %>% ggtsdisplay(lag.max = 168*5)
train.bc %>% diff(lag=24) %>% ggtsdisplay(lag.max = 168)
train.bc %>% diff() %>% diff(lag=24) %>% ggtsdisplay(lag.max = 168)
train.bc %>% diff(lag=168) %>% ggtsdisplay(lag.max = 168*5)
train.bc %>% diff() %>% diff(lag=168) %>% ggtsdisplay(lag.max = 168*5) # arima (0,1,0)(0,1,1)[168]
train.bc %>% diff() %>% diff(lag=168) %>% ggtsdisplay(lag.max = 168) # arima (0,1,0)(0,1,1)[168]
# acf cuts off after p=168, pacf shows peaks each 168
train.bc %>% diff() %>% diff(lag=24) %>% diff(lag=168) %>% ggtsdisplay(lag.max = 168*5)
```

@@

```{r save RData, include=FALSE}
gc()
proj_file <- 'proj.mems.RData'
save.image(proj_file)
```

>>>>

---
title: "Project Membach"
author: "Marco Ciotola"
output:
  pdf_document:
    keep_tex: yes
    fig_width: 6.5
    fig_height: 4
#  html_document:
#    mathjax: local
#    self_contained: no
#    fig_width: 6.5
#    fig_height: 4
---

@@

```{r bc acf-pacf}
train.bc %>% ggtsdisplay(lag.max = 168*5)
train.bc %>% diff() %>% ggtsdisplay(lag.max = 168*5)
train.bc %>% diff(lag=24) %>% ggtsdisplay(lag.max = 168)
train.bc %>% diff() %>% diff(lag=24) %>% ggtsdisplay(lag.max = 168)
train.bc %>% diff(lag=168) %>% ggtsdisplay(lag.max = 168*5)
train.bc %>% diff() %>% diff(lag=168) %>% ggtsdisplay(lag.max = 168*5) # arima (0,1,0)(0,1,1)[168]
train.bc %>% diff() %>% diff(lag=168) %>% ggtsdisplay(lag.max = 168) # arima (0,1,0)(0,1,1)[168]
# acf cuts off after p=168, pacf shows peaks each 168
train.bc %>% diff() %>% diff(lag=24) %>% diff(lag=168) %>% ggtsdisplay(lag.max = 168*5)
```

@@

```{r save RData, include=FALSE}
gc()
save.image(proj_file)
```
